---
title: "Estimating Annual Median Household Income in St. Louis City"
subtitle: "A TWDeConv Example: Deconvolving ACS 5-Year Estimates"
author: "TWDeConv Example"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

# Background

The ACS 5-year estimate for median household income (`B19013_001`) in Census
tract $i$ ending in year $s$ is approximately:

$$Y_{i,s} = \frac{1}{5}\sum_{t=s-4}^{s} X_{i,t} + \varepsilon_{i,s}$$

where $X_{i,t}$ is the **true annual** median income — which is never directly
observed. **TWDeConv** recovers $X$ by solving a Kronecker-structured
spatio-temporal generalised lasso via ADMM.

St. Louis City is an **independent city** in Missouri (county FIPS `510`),
making it a compact, well-defined study area with ~100 Census tracts.

# Setup

```{r packages}
library(TWDeConv)
library(ggplot2)
library(sf)
library(Matrix)
```

> **Census API key required.**  If you have not already registered, get a free
> key at <https://api.census.gov/data/key_signup.html> and install it once with:
> ```r
> tidycensus::census_api_key("YOUR_KEY_HERE", install = TRUE)
> ```

# Fetch and Prepare ACS Data

We use four consecutive ACS 5-year windows (end years 2019–2022). This gives:

- $S = 4$ observed windows
- $T = S + 4 = 8$ latent annual time points (2015–2022)

```{r fetch-data, cache=TRUE}
inp <- prep_acs_for_deconv(
  state     = "MO",
  county    = "St. Louis City",
  year_ends = 2019:2022,
  variables = c(med_inc = "B19013_001"),
  k         = 2L,          # 2nd-order temporal difference (trend filter)
  queen     = TRUE,         # Queen contiguity for spatial Laplacian
  vre       = TRUE,         # Use Variance Replicate Estimates for SE
  cache_dir = "~/acs_cache" # persistent cache — avoids re-downloading
)

print(inp)
```

The `prep_acs_for_deconv()` call automatically builds all five matrices needed
by the solver:

| Matrix | Meaning |
|--------|---------|
| `Y`  | $n \times S$ observed ACS estimates |
| `M`  | $S \times T$ convolution matrix (5-year averaging) |
| `W`  | $(nS) \times (nS)$ diagonal precision (inverse-variance) weights |
| `D`  | $(T-k) \times T$ temporal difference penalty |
| `Ls` | $n \times n$ spatial graph Laplacian |

# Solve the Deconvolution

## Single solve with chosen $\lambda$

We start with a single solve using $\lambda_t = 5$ and $\lambda_s = 0.5$.
These values reflect a stronger temporal penalty (income trends are smoother
than year-to-year noise) and a moderate spatial penalty.

```{r solve}
X_hat <- solve_deconv(
  Y        = inp$Y,
  M        = inp$M,
  W        = inp$W,
  D        = inp$D,
  Ls       = inp$Ls,
  lambda_t = 5,
  lambda_s = 0.5,
  rho      = 1.0,
  verbose  = TRUE
)

cat("Recovered signal dimensions:", nrow(X_hat), "tracts x", ncol(X_hat), "years\n")
cat("Latent year range:",
    inp$metadata$first_latent_year, "to", inp$metadata$last_latent_year, "\n")
```

## Optional: return the full solver object for diagnostics

```{r solver-diagnostics, fig.height=4}
solver <- solve_deconv(
  Y             = inp$Y,
  M             = inp$M,
  W             = inp$W,
  D             = inp$D,
  Ls            = inp$Ls,
  lambda_t      = 5,
  lambda_s      = 0.5,
  rho           = 1.0,
  return_solver = TRUE
)

solver$print()

# Convergence trace
conv_df <- solver$history
ggplot(conv_df, aes(x = iter)) +
  geom_line(aes(y = r_primal, colour = "Primal residual")) +
  geom_line(aes(y = r_dual,   colour = "Dual residual")) +
  scale_y_log10() +
  scale_colour_manual(values = c("Primal residual" = "#2c7bb6",
                                 "Dual residual"   = "#d7191c"),
                      name = NULL) +
  labs(title = "ADMM convergence",
       x = "Iteration", y = "Residual norm (log scale)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Results

## Prepare a long-format data frame

```{r tidy-results}
latent_years <- inp$metadata$first_latent_year:inp$metadata$last_latent_year

# Long format: one row per (tract, year)
X_long <- as.data.frame(X_hat)
colnames(X_long) <- as.character(latent_years)
X_long$GEOID     <- rownames(X_hat)

X_long <- tidyr::pivot_longer(
  X_long,
  cols      = -GEOID,
  names_to  = "year",
  values_to = "med_inc_hat"
)
X_long$year <- as.integer(X_long$year)
```

## Time series: selected tracts

Pick a handful of tracts to visualise the recovered annual signal alongside
the observed 5-year ACS values.

```{r time-series, fig.height=5}
# Observed Y in long format
Y_long <- as.data.frame(inp$Y)
colnames(Y_long) <- as.character(2019:2022)
Y_long$GEOID     <- rownames(inp$Y)

Y_long <- tidyr::pivot_longer(
  Y_long,
  cols      = -GEOID,
  names_to  = "year_end",
  values_to = "med_inc_obs"
)
Y_long$year_end <- as.integer(Y_long$year_end)

# Select a few representative tracts (highest, median, lowest income)
inc_summary <- rowMeans(inp$Y, na.rm = TRUE)
sel_geoids  <- names(sort(inc_summary))[
  round(seq(1, length(inc_summary), length.out = 5))
]

X_sel <- X_long[X_long$GEOID %in% sel_geoids, ]
Y_sel <- Y_long[Y_long$GEOID %in% sel_geoids, ]

ggplot() +
  # Observed 5-year MA: horizontal segments spanning the window
  geom_segment(
    data = Y_sel,
    aes(x = year_end - 4, xend = year_end,
        y = med_inc_obs,  yend = med_inc_obs,
        colour = GEOID),
    linewidth = 2.5, alpha = 0.4
  ) +
  # Recovered annual signal
  geom_line(
    data = X_sel,
    aes(x = year, y = med_inc_hat, colour = GEOID),
    linewidth = 0.8
  ) +
  geom_point(
    data = X_sel,
    aes(x = year, y = med_inc_hat, colour = GEOID),
    size = 2
  ) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  scale_x_continuous(breaks = latent_years) +
  labs(
    title    = "Recovered annual vs. observed 5-year median household income",
    subtitle = "Thick segments = ACS 5-year observed windows; lines = recovered annual signal",
    x = "Year", y = "Median household income",
    colour = "Census tract (GEOID)"
  ) +
  theme_bw() +
  theme(legend.position  = "bottom",
        axis.text.x      = element_text(angle = 45, hjust = 1),
        legend.text      = element_text(size = 7))
```

## Spatial maps: recovered income by year

```{r maps, fig.height=8}
# Join X_hat back to sf geometries from the first ACS fetch
# We need to re-fetch with geometry for the sf object
# (prep_acs_for_deconv stores the sf in the first ACS call internally;
#  here we reconstruct from the GEOID order in inp$Y)
# If you saved the sf object separately, use that instead.

# Get tract geometries via tidycensus
tracts_sf <- tidycensus::get_acs(
  geography = "tract",
  variables = "B19013_001",
  state     = "MO",
  county    = "St. Louis City",
  year      = 2022,
  survey    = "acs5",
  geometry  = TRUE
) |>
  dplyr::select(GEOID, geometry) |>
  dplyr::distinct(GEOID, .keep_all = TRUE)

# Attach recovered estimates for selected years
years_to_map <- c(2015, 2017, 2019, 2021, 2022)

map_df <- X_long |>
  dplyr::filter(year %in% years_to_map) |>
  dplyr::left_join(tracts_sf, by = "GEOID") |>
  sf::st_as_sf()

ggplot(map_df) +
  geom_sf(aes(fill = med_inc_hat / 1000), colour = NA) +
  facet_wrap(~ year, ncol = 3) +
  scale_fill_viridis_c(
    option = "magma", direction = -1,
    name   = "Median income\n($000s)",
    na.value = "grey80"
  ) +
  labs(
    title    = "Recovered annual median household income — St. Louis City tracts",
    subtitle = paste0("TWDeConv deconvolution  |  λ_t = 5,  λ_s = 0.5",
                      "  |  Latent years ", min(latent_years), "–", max(latent_years))
  ) +
  theme_void() +
  theme(
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "right",
    plot.title       = element_text(face = "bold"),
    plot.subtitle    = element_text(colour = "grey40", size = 9)
  )
```

## Summary statistics across tracts

```{r summary-stats}
# Compare observed (Y) vs. recovered (X_hat) distributions
cat("--- Observed ACS 5-year estimates (Y) ---\n")
print(summary(as.vector(inp$Y)))

cat("\n--- Recovered annual estimates (X_hat) ---\n")
print(summary(as.vector(X_hat)))

# Coefficient of variation across years (how much annual variation recovered)
cv_tract <- apply(X_hat, 1, function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
cat("\n--- Within-tract CV of recovered annual signal ---\n")
print(summary(cv_tract))
```

# Sensitivity to $\lambda$ Parameters

Since we lack ground truth for real data, it is good practice to check how
sensitive the results are to the choice of regularization parameters.

```{r lambda-sensitivity, fig.height=4}
lambdas <- c(1, 5, 10, 20)

results_lt <- lapply(lambdas, function(lt) {
  X_tmp <- solve_deconv(
    Y = inp$Y, M = inp$M, W = inp$W, D = inp$D, Ls = inp$Ls,
    lambda_t = lt, lambda_s = 0.5
  )
  data.frame(
    lambda_t = lt,
    year     = latent_years,
    mean_inc = colMeans(X_tmp, na.rm = TRUE)
  )
})
results_lt <- do.call(rbind, results_lt)
results_lt$lambda_t <- factor(results_lt$lambda_t)

ggplot(results_lt, aes(x = year, y = mean_inc / 1000,
                        colour = lambda_t, group = lambda_t)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1, suffix = "k")) +
  scale_colour_brewer(palette = "RdYlBu", name = expression(lambda[t])) +
  labs(
    title = expression("Sensitivity to " * lambda[t] * "  (city-wide mean income)"),
    subtitle = expression(lambda[s] * " = 0.5  fixed"),
    x = "Year", y = "Mean median household income ($000s)"
  ) +
  theme_bw() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

# Session Info

```{r session-info}
sessionInfo()
```
