---
title: "Estimating Annual Non-Hispanic Black Population Share in Cook County, IL"
subtitle: "A TWDeConv Example: Deconvolving ACS 5-Year Estimates and Validating Against the 2020 Decennial Census"
author: "TWDeConv Example"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

# Background

The ACS 5-year estimate for the proportion of non-Hispanic (NH) Black residents
in Census tract $i$ ending in year $s$ is approximately:

$$P_{i,s} = \frac{1}{5}\sum_{t=s-4}^{s} X_{i,t} + \varepsilon_{i,s}$$

where $X_{i,t}$ is the **true annual** NH Black population share — never
directly observed at the annual, tract level.  **TWDeConv** recovers $X$ by
solving a Kronecker-structured spatio-temporal generalised lasso via ADMM.

Because the observable $P_{i,s}$ is a **derived proportion** (not a single
ACS variable), we compute it and its standard error from two ACS variables:

| Variable | Description |
|----------|-------------|
| `B03002_001` | Total population |
| `B03002_004` | Not Hispanic or Latino: Black or African American alone |

$$P_{i,s} = \frac{\hat{X}^{\text{Black}}_{i,s}}{\hat{X}^{\text{Total}}_{i,s}},
\quad
\widehat{\text{SE}}(P_{i,s}) = \frac{1}{\hat{X}^{\text{Total}}_{i,s}}
\sqrt{\widehat{\text{SE}}^2_{\text{Black}} - P_{i,s}^2\,
\widehat{\text{SE}}^2_{\text{Total}}}$$

(ACS ratio SE formula; falls back to the sum-of-squares form when the
discriminant is negative.)

The recovered 2020 annual estimates are then validated against the
**2020 Decennial Census** PL 94-171 redistricting file — a full enumeration
with no sampling error — making this the only example in the package with
near-true ground truth at the tract level.

Cook County, IL contains ~1,300 Census tracts, a substantially larger
problem than the St. Louis City income example.

# Setup

```{r packages}
library(TWDeConv)
library(ggplot2)
library(sf)
library(Matrix)

options(tigris_use_cache = TRUE)
```

> **Census API key required.**  Register at
> <https://api.census.gov/data/key_signup.html> and install once with:
> ```r
> tidycensus::census_api_key("YOUR_KEY_HERE", install = TRUE)
> ```

# Fetch and Prepare ACS Data

We use five consecutive ACS 5-year windows (end years 2018–2022):

- $S = 5$ observed windows
- $T = S + 4 = 9$ latent annual time points (2014–2022)

## Step 1: Structural matrices from NH Black counts

`prep_acs_for_deconv()` is called with `B03002_004` (NH Black count) to
obtain the five structural matrices ($Y$, $M$, $W$, $D$, $L_s$) and the
aligned GEOID set.  We will replace $Y$ and $W$ with proportion-based
counterparts in Step 2.

```{r fetch-black, cache=TRUE}
year_ends <- 2018:2022
S         <- length(year_ends)

inp_black <- prep_acs_for_deconv(
  state     = "IL",
  county    = "Cook",
  year_ends = year_ends,
  variables = c(black_nh = "B03002_004"),
  k         = 2L,
  queen     = TRUE,
  vre       = TRUE,
  cache_dir = "~/acs_cache"
)
print(inp_black)

common_geoids <- rownames(inp_black$Y)
n_tracts      <- length(common_geoids)
cat("Common tracts across all windows:", n_tracts, "\n")
```

## Step 2: Fetch total population (denominator)

```{r fetch-total, cache=TRUE}
total_list <- vector("list", S)
for (i in seq_len(S)) {
  total_list[[i]] <- fetch_acs_vre_data(
    state     = "IL",
    county    = "Cook",
    year_end  = year_ends[i],
    variables = c(total = "B03002_001"),
    geometry  = FALSE,
    vre       = FALSE,          # VRE not needed for denominator
    cache_dir = "~/acs_cache",
    quiet     = TRUE
  )
}
```

## Step 3: Compute proportion matrix Y and precision weights W

```{r build-rates}
# -- NH Black counts and their SEs (n x S) ----------------------------------
Y_count  <- inp_black$Y
SE_count <- matrix(
  sqrt(1 / Matrix::diag(inp_black$W)),
  nrow = n_tracts, ncol = S,
  dimnames = list(common_geoids, colnames(Y_count))
)

# -- Total population and its SE (n x S), aligned to common_geoids ----------
denom_mat    <- matrix(NA_real_, nrow = n_tracts, ncol = S,
                       dimnames = list(common_geoids, colnames(Y_count)))
SE_denom_mat <- denom_mat

for (i in seq_len(S)) {
  g_i <- rownames(total_list[[i]]$estimates)
  idx <- match(common_geoids, g_i)
  denom_mat[, i]    <- total_list[[i]]$estimates[idx, "total"]
  SE_denom_mat[, i] <- total_list[[i]]$se[idx, "total"]
}

# -- Proportion: P = NH Black / Total ---------------------------------------
Y_prop <- Y_count / denom_mat

# -- SE of proportion (ACS ratio formula) ------------------------------------
SE_prop <- matrix(NA_real_, nrow = n_tracts, ncol = S,
                  dimnames = dimnames(Y_prop))

for (i in seq_len(S)) {
  N    <- denom_mat[, i]
  p    <- Y_prop[, i]
  se_x <- SE_count[, i]
  se_n <- SE_denom_mat[, i]

  disc        <- se_x^2 - p^2 * se_n^2
  SE_prop[, i] <- ifelse(
    !is.na(disc) & disc >= 0,
    sqrt(disc)                      / N,   # standard ACS ratio formula
    sqrt(se_x^2 + p^2 * se_n^2)    / N    # fallback (sum form)
  )
}

# Guard: set SE to NA where total is zero or missing
SE_prop[is.na(denom_mat) | denom_mat == 0] <- NA_real_

cat("Y_prop range  :", round(range(Y_prop, na.rm = TRUE), 3), "\n")
cat("SE_prop range :", round(range(SE_prop, na.rm = TRUE), 5), "\n")

# Precision weight matrix (NA SEs produce W=0, excluding from WLS)
W_prop <- build_precision_weights(SE_prop)
cat("W diagonal range:", round(range(Matrix::diag(W_prop)), 1), "\n")
```

# Solve the Deconvolution

Proportions live on $[0, 1]$, so $\lambda$ values are much smaller than for
dollar-denominated income.  We use $\lambda_t = 1$ and $\lambda_s = 0.1$.

> **Note:** With ~1,300 Cook County tracts the one-time Cholesky factorisation
> in `$setup()` may take 1–2 minutes.

```{r solve, cache=TRUE}
X_hat <- solve_deconv(
  Y        = Y_prop,
  M        = inp_black$M,
  W        = W_prop,
  D        = inp_black$D,
  Ls       = inp_black$Ls,
  lambda_t = 1,
  lambda_s = 0.1,
  rho      = 1.0,
  max_iter = 5000,
  verbose  = TRUE
)

# Clip to [0, 1]: proportions must be bounded
X_hat <- pmax(pmin(X_hat, 1), 0)

cat("Recovered dimensions:", nrow(X_hat), "tracts x", ncol(X_hat), "years\n")
cat("X_hat range:", round(range(X_hat, na.rm = TRUE), 3), "\n")
cat("Latent years:",
    inp_black$metadata$first_latent_year, "–",
    inp_black$metadata$last_latent_year, "\n")
```

## Convergence diagnostic

```{r solver-diagnostics, fig.height=4}
solver <- solve_deconv(
  Y             = Y_prop,
  M             = inp_black$M,
  W             = W_prop,
  D             = inp_black$D,
  Ls            = inp_black$Ls,
  lambda_t      = 1,
  lambda_s      = 0.1,
  rho           = 1.0,
  max_iter      = 5000,
  return_solver = TRUE
)
solver$print()

ggplot(solver$history, aes(x = iter)) +
  geom_line(aes(y = r_primal, colour = "Primal residual")) +
  geom_line(aes(y = r_dual,   colour = "Dual residual")) +
  scale_y_log10() +
  scale_colour_manual(
    values = c("Primal residual" = "#2c7bb6", "Dual residual" = "#d7191c"),
    name   = NULL
  ) +
  labs(title = "ADMM convergence",
       x = "Iteration", y = "Residual norm (log scale)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Results

## Long-format data frame

```{r tidy-results}
latent_years <- inp_black$metadata$first_latent_year:inp_black$metadata$last_latent_year

X_long <- as.data.frame(X_hat)
colnames(X_long) <- as.character(latent_years)
X_long$GEOID     <- rownames(X_hat)

X_long <- tidyr::pivot_longer(
  X_long,
  cols      = -GEOID,
  names_to  = "year",
  values_to = "prop_black_hat"
)
X_long$year <- as.integer(X_long$year)
```

## Time series: selected tracts

```{r time-series, fig.height=5}
# Observed proportions in long format
Y_long <- as.data.frame(Y_prop)
colnames(Y_long) <- as.character(year_ends)
Y_long$GEOID     <- rownames(Y_prop)

Y_long <- tidyr::pivot_longer(
  Y_long,
  cols      = -GEOID,
  names_to  = "year_end",
  values_to = "prop_obs"
)
Y_long$year_end <- as.integer(Y_long$year_end)

# Select 5 representative tracts spanning the NH Black share distribution
mean_prop  <- rowMeans(Y_prop, na.rm = TRUE)
sel_geoids <- names(sort(mean_prop))[
  round(seq(1, sum(!is.na(mean_prop)), length.out = 5))
]

X_sel <- X_long[X_long$GEOID %in% sel_geoids, ]
Y_sel <- Y_long[Y_long$GEOID %in% sel_geoids, ]

ggplot() +
  geom_segment(
    data = Y_sel,
    aes(x = year_end - 4, xend = year_end,
        y = prop_obs, yend = prop_obs, colour = GEOID),
    linewidth = 2.5, alpha = 0.4
  ) +
  geom_line(
    data = X_sel,
    aes(x = year, y = prop_black_hat, colour = GEOID),
    linewidth = 0.8
  ) +
  geom_point(
    data = X_sel,
    aes(x = year, y = prop_black_hat, colour = GEOID),
    size = 2
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = latent_years) +
  labs(
    title    = "Recovered annual vs. observed 5-year NH Black population share",
    subtitle = "Thick segments = ACS 5-year observed windows; lines = recovered annual signal",
    x = "Year", y = "NH Black share",
    colour = "Census tract (GEOID)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x     = element_text(angle = 45, hjust = 1),
        legend.text     = element_text(size = 7))
```

## Spatial maps: recovered share by year

```{r maps, fig.height=8}
tracts_sf <- tidycensus::get_acs(
  geography = "tract",
  variables = "B03002_001",
  state     = "IL",
  county    = "Cook",
  year      = 2022,
  survey    = "acs5",
  geometry  = TRUE
) |>
  dplyr::select(GEOID, geometry) |>
  dplyr::distinct(GEOID, .keep_all = TRUE)

years_to_map <- c(2014, 2016, 2018, 2020, 2022)

map_df <- X_long |>
  dplyr::filter(year %in% years_to_map) |>
  dplyr::left_join(tracts_sf, by = "GEOID") |>
  sf::st_as_sf()

ggplot(map_df) +
  geom_sf(aes(fill = prop_black_hat), colour = NA) +
  facet_wrap(~ year, ncol = 3) +
  scale_fill_viridis_c(
    option   = "magma", direction = -1,
    name     = "NH Black\nshare",
    labels   = scales::percent_format(accuracy = 1),
    na.value = "grey80"
  ) +
  labs(
    title    = "Recovered annual NH Black population share — Cook County tracts",
    subtitle = paste0("TWDeConv  |  \u03bb_t = 1,  \u03bb_s = 0.1  |  Latent years ",
                      min(latent_years), "\u2013", max(latent_years))
  ) +
  theme_void() +
  theme(
    strip.text      = element_text(face = "bold", size = 10),
    legend.position = "right",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(colour = "grey40", size = 9)
  )
```

# Validation: Comparison with 2020 Decennial Census

## Why the Decennial Census works here

Unlike income (which the Decennial does not collect), **racial composition is
measured by the Decennial Census** as a full enumeration with no sampling
error.  The 2020 PL 94-171 redistricting file provides:

| Variable | Description |
|----------|-------------|
| `P2_001N` | Total population |
| `P2_006N` | Not Hispanic or Latino: Black or African American alone |

Two caveats apply:

1. **Timing**: the Decennial is a single-day snapshot (April 1, 2020); the
   deconvolved estimate represents an annual average.  In rapidly changing
   tracts these may differ.
2. **Mode effects**: mail/online survey (ACS) vs. enumerator (Decennial) can
   produce small differences in self-reported race.

## Fetch Decennial 2020 tract data

```{r decennial-2020, cache=TRUE}
dec_raw <- tidycensus::get_decennial(
  geography = "tract",
  variables = c(total    = "P2_001N",
                black_nh = "P2_006N"),
  state     = "IL",
  county    = "Cook",
  year      = 2020,
  sumfile   = "pl"
)

dec_2020 <- dec_raw |>
  dplyr::select(GEOID, variable, value) |>
  tidyr::pivot_wider(names_from = variable, values_from = value) |>
  dplyr::mutate(
    prop_black_dec = dplyr::if_else(total > 0, black_nh / total, NA_real_)
  )

cat("Decennial tracts:", nrow(dec_2020), "\n")
cat("Proportion range:", round(range(dec_2020$prop_black_dec, na.rm = TRUE), 3), "\n")
```

## Agreement statistics

```{r comparison}
comp_df <- X_long |>
  dplyr::filter(year == 2020) |>
  dplyr::inner_join(dec_2020, by = "GEOID") |>
  dplyr::filter(!is.na(prop_black_hat), !is.na(prop_black_dec))

rmse <- sqrt(mean((comp_df$prop_black_hat - comp_df$prop_black_dec)^2))
mae  <- mean(abs(comp_df$prop_black_hat  - comp_df$prop_black_dec))
bias <- mean(comp_df$prop_black_hat      - comp_df$prop_black_dec)
r    <- cor(comp_df$prop_black_hat, comp_df$prop_black_dec)

cat("=== TWDeConv 2020 vs. Decennial 2020 (Cook County tracts) ===\n")
cat(sprintf("  Tracts compared : %d\n",      nrow(comp_df)))
cat(sprintf("  Pearson r       : %.4f\n",    r))
cat(sprintf("  RMSE            : %.4f  (%.1f pct pts)\n", rmse, rmse * 100))
cat(sprintf("  MAE             : %.4f  (%.1f pct pts)\n", mae,  mae  * 100))
cat(sprintf("  Mean bias       : %+.4f (%+.1f pct pts)\n", bias, bias * 100))
```

## Scatter plot: TWDeConv vs. Decennial

```{r scatter, fig.height=5}
ggplot(comp_df, aes(x = prop_black_dec, y = prop_black_hat)) +
  geom_point(alpha = 0.35, size = 1.2, colour = "#2c7bb6") +
  geom_abline(slope = 1, intercept = 0,
              colour = "black", linetype = "dashed", linewidth = 0.7) +
  geom_smooth(method = "lm", colour = "#d7191c",
              se = FALSE, linewidth = 0.9) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  annotate("text", x = 0.05, y = 0.92, hjust = 0, size = 3.8,
           label = sprintf("r = %.3f\nRMSE = %.1f pct pts\nBias = %+.1f pct pts",
                           r, rmse * 100, bias * 100)) +
  labs(
    title    = "Deconvolved 2020 NH Black share vs. 2020 Decennial Census",
    subtitle = "Dashed = perfect agreement  |  Red = OLS fit  |  Each point = one Census tract",
    x        = "Decennial Census 2020 (full count)",
    y        = "TWDeConv deconvolved 2020"
  ) +
  theme_bw()
```

## Side-by-side maps: TWDeConv 2020 vs. Decennial 2020

```{r map-comparison, fig.height=6}
map_comp <- comp_df |>
  dplyr::select(GEOID, prop_black_hat, prop_black_dec) |>
  dplyr::left_join(tracts_sf, by = "GEOID") |>
  sf::st_as_sf() |>
  tidyr::pivot_longer(
    cols      = c(prop_black_hat, prop_black_dec),
    names_to  = "source",
    values_to = "prop_black"
  ) |>
  dplyr::mutate(source = dplyr::recode(source,
    prop_black_hat = "TWDeConv 2020",
    prop_black_dec = "Decennial Census 2020"
  ))

ggplot(map_comp) +
  geom_sf(aes(fill = prop_black), colour = NA) +
  facet_wrap(~ source) +
  scale_fill_viridis_c(
    option   = "magma", direction = -1,
    name     = "NH Black\nshare",
    labels   = scales::percent_format(accuracy = 1),
    na.value = "grey80"
  ) +
  labs(
    title    = "NH Black population share: TWDeConv 2020 vs. Decennial Census 2020",
    subtitle = "Cook County Census tracts"
  ) +
  theme_void() +
  theme(
    strip.text      = element_text(face = "bold", size = 11),
    legend.position = "right",
    plot.title      = element_text(face = "bold")
  )
```

## Residual map: TWDeConv − Decennial

```{r map-residual, fig.height=5}
map_resid <- comp_df |>
  dplyr::mutate(resid = prop_black_hat - prop_black_dec) |>
  dplyr::left_join(tracts_sf, by = "GEOID") |>
  sf::st_as_sf()

lim <- max(abs(map_resid$resid), na.rm = TRUE)

ggplot(map_resid) +
  geom_sf(aes(fill = resid), colour = NA) +
  scale_fill_distiller(
    palette  = "RdBu",
    limits   = c(-lim, lim),
    name     = "TWDeConv \u2212\nDecennial",
    labels   = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Residual: TWDeConv 2020 minus Decennial 2020 NH Black share",
    subtitle = "Blue = underestimate; Red = overestimate"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(colour = "grey40", size = 9)
  )
```

# Session Info

```{r session-info}
sessionInfo()
```
